---
title: "EDA"
author: "Phoebe Abramowitz and Jack Moorer"
date: "4/12/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Data Cleaning

```{r clean data}
#load packages
library(dplyr)
library(astsa)
library(tseries)
library(TSA)
library(tsoutliers)
```

```{r}
#read in data
raw_birthrates = read.csv("../data/daily-total-female-births-in-cal.csv")

#remove final row
clean_birthrates <- filter(raw_birthrates, Date!="Daily total female births in California")

#define column names
colnames(clean_birthrates) <-  c('date',"daily_female_births")

#convert dates to Data data type
clean_birthrates$date <- as.Date(clean_birthrates$date)

#make dates row names
birthrates <- clean_birthrates$daily_female_births
female_birthrates <- data.frame(birthrates)
rownames(female_birthrates) <- clean_birthrates$date

#examine data
head(female_birthrates)
```

```{r}
#convert to time series
female_birthrates <- ts(female_birthrates)
```

#EDA
First let's check for outliers using tso

```{r}
#check for outliers
outliers <- tso(female_birthrates, types = c("TC", "AO", "LS", "IO", "SLS"))
outliers
```

It looks like an Additive Outlier, or isolated speak, is present at day 266.

Let's look at the plot.

```{r}
#plot outliers
plot(outliers)
```

As you can see we have one outlier affecting our data.

```{r}
#get index and time of outlier
outlier_indx <- outliers$outliers$ind

#length of ts
n <- length(female_birthrates)

#find outlier effect
ao <- outliers("AO", outlier_indx)
ao_effect <- outliers.effects(ao, n)
coefhat <- as.numeric(outliers$outliers['coefhat'])
ao_effect <- coefhat*ao_effect

#generate outlier affect time series
ao_effect_ts <- ts(ao_effect, frequency = frequency(female_birthrates), start= start(female_birthrates))

#generate time series without outlier
birthrates_wo_ao <- female_birthrates - ao_effect_ts

#plot original time series and time series without outlier
par(mfrow = c(3, 1))

plot(female_birthrates, main = "Original Time Series")

plot(birthrates_wo_ao, main = "Time Series without Outlier")

plot(ao_effect_ts, main = "Outlier Plot")
```

Let's just remove this outlier and continue our analysis.

```{r}
#generate birthrates time series without outlier
rownames(birthrates_wo_ao) <- clean_birthrates$date

#replace female_birthrates data frame
female_birthrates <- birthrates_wo_ao
```

```{r}
#plot time series
plot.ts(female_birthrates, main = "Time Series Plot of Female Birthrate Data")
```

Let's plot the ACF and PACF of this time series.

```{r}
#plot acf of ts
acf(female_birthrates, main = "ACF of Birthrates Time Series")
```

```{r}
#plot pacf
pacf(female_birthrates, main = "PACF of Female Birthrates")
```

It appears our data is not stationary. One issue might be there is an underlying trend in the data, and the mean is dependent on time.


Let's see if there is an underlying trend in our data by fitting a linear model.

```{r}
#plot ts with regression line
plot.ts(female_birthrates, main = "Time Series with OLS regression line")
abline(reg = lm(female_birthrates ~ time(female_birthrates)), col = 'red')
```
 
It seems like there is a slight trend. Let's plot the residuals of the regression, which should be the detrended time series.


```{r}
#plot detrended time series vie OLS regression
plot(resid(reg), type = 'l', main = "Detrended via Simple OLS", ylab = "Birthrates Detrended", xlab = "Time")
abline(lm(resid(reg) ~ time(resid(reg))), col = 'red')
```
 
We can see now our data is detrended, and it appears the mean of the time series in not dependent on time.

Let's look at the acf and pacf of the detrended via OLS time series.

```{r}
#acf of time series detrended via OLS
acf(reg$residuals, main = "ACF of Birthrates Detrended via OLS")
```
 
```{r}
#pacf of time series detrended via OLS
pacf(reg$residuals, main = "PACF of Birthrates Detrended via OLS")
```

While detrended via OLS looks like it produces a detrended time series, the ACF and PACF do not seem to be tailing off, which implies the time series may not be stationary.
 
Let's take a look at the first difference of the time series.
 
```{r}
#take the first difference of the time series
diff_vals <- diff(female_birthrates)

#plot diff of time series with linear region line
ts.plot(diff(female_birthrates), main = "First Difference of Birthrates Time Series")
abline(reg = lm(diff(female_birthrates) ~ time(diff(female_birthrates))), col = 'red')
```
 
It seems like the first difference does not have an underlying trend, which is expected.

Let's look at the ACF and PACF of the first difference to see if this time series is stationary.

```{r}
#plot acf of first diffence
acf(first_diff, main = "ACF of First Difference of Birthrates")
```
 
```{r}
#pacf of first difference of time series
pacf(first_diff, main = "PACF of First Difference of Birthrates")
```

This seems better than the time series detrended via OLS, but still does not look great.

Let's look at the second difference.

```{r}
#get second diff
second_diff_vals <- diff(diff(female_birthrates))

#plot second diff of time series with linear region line
ts.plot(diff(diff(female_birthrates)), main = "First Difference of Birthrates Time Series")
abline(reg = lm(diff(diff(female_birthrates)) ~ time(diff(diff(female_birthrates)))), col = 'red')
```


```{r}
#acf of second diff
acf(second_diff, main = "ACF of Second Difference of Birthrates")
```

```{r}
#plot pacf
pacf(second_diff, main = "PACF of Second Difference of Birthrates")
```

We want to avoid overdifferencing, so we only want to use the second difference if we can tell the first difference is still not stationary. One way to check that is by using a Dickey-Fuller test.


```{r}
adf.test(female_birthrates, alternative = "stationary", k = 6)
adf.test(female_birthrates, alternative = "stationary", k = 3)
adf.test(female_birthrates, alternative = "explosive", k = 0)
```

```{r}
adf.test(diff(female_birthrates), alternative = "stationary")
adf.test(diff(female_birthrates), alternative = "explosive", k = 0)
```


There is also a good chance that we have seasonality here.

=======

```{r}
#convert to time series
female_birthrates <- ts(female_birthrates)
```

#EDA
First let's check for outliers using tso

```{r}
#check for outliers
outliers <- tso(female_birthrates, types = c("TC", "AO", "LS", "IO", "SLS"))
outliers
```

It looks like an Additive Outlier, or isolated speak, is present at day 266.

Let's look at the plot.

```{r}
#plot outliers
plot(outliers)
```

As you can see we have one outlier affecting our data.

```{r}
#get index and time of outlier
outlier_indx <- outliers$outliers$ind

#length of ts
n <- length(female_birthrates)

#find outlier effect
ao <- outliers("AO", outlier_indx)
ao_effect <- outliers.effects(ao, n)
coefhat <- as.numeric(outliers$outliers['coefhat'])
ao_effect <- coefhat*ao_effect

#generate outlier affect time series
ao_effect_ts <- ts(ao_effect, frequency = frequency(female_birthrates), start= start(female_birthrates))

#generate time series without outlier
birthrates_wo_ao <- female_birthrates - ao_effect_ts

#plot original time series and time series without outlier
par(mfrow = c(3, 1))

plot(female_birthrates, main = "Original Time Series")

plot(birthrates_wo_ao, main = "Time Series without Outlier")

plot(ao_effect_ts, main = "Outlier Plot")
```

Let's just remove this outlier and continue our analysis.

```{r}
#generate birthrates time series without outlier
rownames(birthrates_wo_ao) <- clean_birthrates$date

#replace female_birthrates data frame
female_birthrates <- birthrates_wo_ao
```

```{r}
#plot time series
plot.ts(female_birthrates, main = "Time Series Plot of Female Birthrate Data")
```

Let's plot the ACF and PACF of this time series.

```{r}
#plot acf of ts
acf(female_birthrates, main = "ACF of Birthrates Time Series")
```

```{r}
#plot pacf
pacf(female_birthrates, main = "PACF of Female Birthrates")
```

It appears our data is not stationary. One issue might be there is an underlying trend in the data, and the mean is dependent on time.


Let's see if there is an underlying trend in our data by fitting a linear model.

```{r}
#plot ts with regression line
plot.ts(female_birthrates, main = "Time Series with OLS regression line")
abline(reg = lm(female_birthrates ~ time(female_birthrates)), col = 'red')
```
 
It seems like there is a slight trend. Let's plot the residuals of the regression, which should be the detrended time series.


```{r}
#plot detrended time series vie OLS regression
plot(resid(reg), type = 'l', main = "Detrended via Simple OLS", ylab = "Birthrates Detrended", xlab = "Time")
abline(lm(resid(reg) ~ time(resid(reg))), col = 'red')
```
 
We can see now our data is detrended, and it appears the mean of the time series in not dependent on time.

Let's look at the acf and pacf of the detrended via OLS time series.

```{r}
#acf of time series detrended via OLS
acf(reg$residuals, main = "ACF of Birthrates Detrended via OLS")
```
 
```{r}
#pacf of time series detrended via OLS
pacf(reg$residuals, main = "PACF of Birthrates Detrended via OLS")
```

While detrended via OLS looks like it produces a detrended time series, the ACF and PACF do not seem to be tailing off, which implies the time series may not be stationary.
 
Let's take a look at the first difference of the time series.
 
```{r}
#take the first difference of the time series
diff_vals <- diff(female_birthrates)

#plot diff of time series with linear region line
ts.plot(diff(female_birthrates), main = "First Difference of Birthrates Time Series")
abline(reg = lm(diff(female_birthrates) ~ time(diff(female_birthrates))), col = 'red')
```
 
It seems like the first difference does not have an underlying trend, which is expected.

Let's look at the ACF and PACF of the first difference to see if this time series is stationary.

```{r}
#plot acf of first diffence
acf(first_diff, main = "ACF of First Difference of Birthrates")
```
 
```{r}
#pacf of first difference of time series
pacf(first_diff, main = "PACF of First Difference of Birthrates")
```

This seems better than the time series detrended via OLS, but still does not look great.

Let's look at the second difference.

```{r}
#get second diff
second_diff_vals <- diff(diff(female_birthrates))

#plot second diff of time series with linear region line
ts.plot(diff(diff(female_birthrates)), main = "First Difference of Birthrates Time Series")
abline(reg = lm(diff(diff(female_birthrates)) ~ time(diff(diff(female_birthrates)))), col = 'red')
```


```{r}
#acf of second diff
acf(second_diff, main = "ACF of Second Difference of Birthrates")
```

```{r}
#plot pacf
pacf(second_diff, main = "PACF of Second Difference of Birthrates")
```

We want to avoid overdifferencing, so we only want to use the second difference if we can tell the first difference is still not stationary. One way to check that is by using a Dickey-Fuller test.


```{r}
adf.test(female_birthrates, alternative = "stationary", k = 6)
adf.test(female_birthrates, alternative = "stationary", k = 3)
adf.test(female_birthrates, alternative = "explosive", k = 0)
```

```{r}
adf.test(diff(female_birthrates), alternative = "stationary")
adf.test(diff(female_birthrates), alternative = "explosive", k = 0)
```


There is also a good chance that we have seasonality here.



