---
title: "Spectral Analysis"
author: "Phoebe Abramowitz"
date: "4/13/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
#load packages
library(dplyr)
library(astsa)
library(tseries)
library(TSA)
```

```{r}
source('./CleanData.R')
```

Detrend using differencing once. 
```{r}
fem_birth <- diff(female_birthrates)
```


Look at the periodogram, then the smoothed periodogram
```{r}
spec.pgram(fem_birth)
#try different values for the kernel & taper
mvspec(fem_birth,kernel('daniell',3), taper = 0.1)
mvspec(fem_birth,kernel('modified.daniell',3), taper = 0.2)
mvspec(fem_birth,kernel('modified.daniell',5), taper = 0)
#This is the best one that we use going forward
#Note for Jack: I vaguely overheard Jonny say in OH that people usually use log scale, 
#but I'll leave that to your discretion if you wanna take that log arg out 
spec.pgram(fem_birth,kernel('modified.daniell',c(1,6,1)), taper = 0.1,log="no")
```
Note for Jack: Is there better phrasing than "top frequencies"?

From this smoothed periodogram, we can choose the top three lags and feed the modal. We append the first fundamental frequency and exhibit all local maxima on the periodogram before selecting the top three lags. 
```{r}
pgram <- spec.pgram(fem_birth,kernel('modified.daniell',c(1,6,1)), taper = 0.1,log="no")
key_freq_ind <- c(1, which(diff(sign(diff(pgram$spec)))==-2) + 1)
key_freq <- pgram$freq[key_freq_ind]
abline(v=key_freq, lty=3)
top_freq <- key_freq[order(pgram$spec[key_freq_ind], decreasing = T)][1:3]
```
The top three frequencies we've observed are: 
```{r}
top_freq
```


We can then compare with the parametric spectral estimator (the red dotted line):

```{r}
pgram <- spec.pgram(fem_birth,kernel('modified.daniell',c(1,6,1)), taper = 0.1,log="no")
pgram_ar <- spec.ar(fem_birth, plot=F)
lines(pgram_ar$freq, pgram_ar$spec, lty=2, col="red")
```
We can use our three frequencies to generate features (sin terms and cosine terms). We use the following model:\\
$$t=\alpha+\beta t+\sum_{j=1}^3 c_jcos(2\pi \omega*jt)+d_j cos(2\pi \omega*jt)+w_t$$

```{r}
t <- 1:length(female_birthrates)
periodic_terms <- do.call(cbind, lapply(top_freq, function(freq) {
  cbind(cos(2 * pi * freq * t), sin(2 * pi * freq * t))
}))
df <- data.frame(female_birthrates, t, periodic_terms)
fit_final <- lm(female_birthrates ~ ., df)
plot(t, female_birthrates, type="l")
lines(t, fit_final$fitted.values, lty=2, col="red")
```
Then, lastly, we'll forecast the next 30 obervations(1 month) with a 95% confidence level.
```{r}
t_new <- (tail(t, 1) + 1):(tail(t, 1) + 30)
periodic_terms_new <- do.call(cbind, lapply(top_freq, function(freq) {
  cbind(cos(2 * pi * freq * t_new), sin(2 * pi * freq * t_new))
}))
df_new <- data.frame(t_new, periodic_terms_new)
colnames(df_new) <- colnames(df)[-1]
predictions <- predict.lm(fit_final, newdata=df_new,interval="prediction", level=.95)
plot(t, female_birthrates, type="l", xlim=c(0, tail(t_new, 1)))
lines(t, fit_final$fitted.values, lty=2, col="red")
lines(t_new, predictions[, "fit"], col="blue")
matlines(t_new, predictions[, 2:3], col = "purple", lty=3)
```

The bounds of our confidence interval encapsulate most of the observed values from the entire dataset at every point, and thus aren't particularly informative. 



