---
title: "Spectral Analysis"
author: "Phoebe Abramowitz"
date: "4/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
#load packages
library(dplyr)
library(astsa)
library(tseries)
library(TSA)
```
Note to Self: Look to this example for guidance
https://bcourses.berkeley.edu/courses/1468637/files?preview=72935387
```{r}
source('./CleanData.R')
#female_birthrates <- read.csv('../data/clean_data.csv')
```
```{r}
# #detrend using Jack's OLS code(consolidate/change later)
# plot.ts(female_birthrates, main = "Time Series with OLS regression line")
# abline(reg=lm(female_birthrates ~ time(female_birthrates)), col = 'red')
# plot(resid(reg), type = 'l', main = "Detrended via Simple OLS", ylab = "Birthrates Detrended", xlab = "Time")
# abline(lm(resid(reg) ~ time(resid(reg))), col = 'red')
# #use fem_birth for now going forward for figuring out spectral analysis
# fem_birth <- reg$residuals
```
Detrend using differencing once. 
```{r}
fem_birth <- diff(female_birthrates)
```


Look at the periodogram, then the smoothed periodogram
```{r}
#spec.pgram automatically accounts for linear detrend, only use fem_birth specifically if detrending a different way
spec.pgram(fem_birth)
#try different values for the kernel & taper
mvspec(fem_birth,kernel('daniell',3), taper = 0.1)
mvspec(fem_birth,kernel('modified.daniell',3), taper = 0.2)
mvspec(fem_birth,kernel('modified.daniell',5), taper = 0)
#Best One in next chunk, use that
```
From this smoothed periodogram, we can choose the top three lags and feed the modal. We append the first fundamental frequency.
```{r}
pgram <- spec.pgram(fem_birth,kernel('modified.daniell',c(1,6,1)), taper = 0.1,log='no')
key_freq_ind <- c(1, which(diff(sign(diff(pgram$spec)))==-2) + 1)
key_freq <- pgram$freq[key_freq_ind]
abline(v=key_freq, lty=3)
top_freq <- key_freq[order(pgram$spec[key_freq_ind], decreasing = T)][1:3]
```

We can then compare with the parametric spectral estimator (the red dotted line):

```{r}
pgram <- spec.pgram(fem_birth,kernel('modified.daniell',c(1,6,1)), taper = 0.1)
#This pse absolutely does not look right
pgram_ar <- spec.ar(fem_birth, plot=F)
lines(pgram_ar$freq, pgram_ar$spec, lty=2, col="red")
```
Use the three frequencies to generate features (sin terms and cosine terms).We use the following model:\\
$$t=\alpha+\beta t+\sum_{j=1}^3 c_jcos(2\pi \omega*jt)+d_j cos(2\pi \omega*jt)+w_t$$

```{r}
t <- 1:length(female_birthrates)
periodic_terms <- do.call(cbind, lapply(top_freq, function(freq) {
  cbind(cos(2 * pi * freq * t), sin(2 * pi * freq * t))
}))
df <- data.frame(female_birthrates, t, periodic_terms)
fit_final <- lm(female_birthrates ~ ., df)
plot(t, female_birthrates, type="l")
lines(t, fit_final$fitted.values, lty=2, col="red")
```
Then, forecast the next 30 obervations(1 month).
```{r}
t_new <- (tail(t, 1) + 1):(tail(t, 1) + 30)
periodic_terms_new <- do.call(cbind, lapply(top_freq, function(freq) {
  cbind(cos(2 * pi * freq * t_new), sin(2 * pi * freq * t_new))
}))
df_new <- data.frame(t_new, periodic_terms_new)
colnames(df_new) <- colnames(df)[-1]
predictions <- predict.lm(fit_final, newdata=df_new,interval="prediction", level=.95)
plot(t, female_birthrates, type="l", xlim=c(0, tail(t_new, 1)))
lines(t, fit_final$fitted.values, lty=2, col="red")
lines(t_new, predictions[, "fit"], col="blue")
matlines(t_new, predictions[, 2:3], col = "purple", lty=3)
```




